---
title: "Cross-species analysis"
author: "Jeremy Miller"
date: "September 15, 2021"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This script performs some statistical analyses and makes some UMAP plots showing glutamatergic neurons in supragranular layers of human MTG and mouse VISp.  One purpose is to quantify to what extent the cell type diversity, variation, and specialization of human types extends beyond that of mouse.  Note that this script is only considering dissociated cells/nuclei (called "FACS" here) and includes code for generating panels of Figure 1.   
If needed set the working directory:  (e.g., `setwd("PATH_TO_DIRECTORY/patchseq_human_L23")`). 
  
### Work space preparation
  
First, load all the libraries (required for all code documents) and set some options.  
  
```{r load libraries, warning=FALSE}
# Load all packages
suppressPackageStartupMessages({
  library(genesorteR)
  library(feather)
  library(dplyr)
  library(gplots)
  library(Seurat)
  library(ggplot2)
  library(cowplot)
  library(scrattch.hicat)
  library(scrattch.vis)
  library(future)
  library(ggbeeswarm)
  library(mfishtools) # getBetaScore
  library(pheatmap)
  library(VENcelltypes) # mito_genes and sex_genes
  library(WGCNA)
  library(Matrix)
  library(gridExtra)
})
options(stringsAsFactors=FALSE)
options(future.globals.maxSize = 4000 * 1024^2)
```
  
  
Next, load the all the relevant data, which were calculated in previous scripts.  
  
```{r load data}
load("input_data_sets.RData")
```
  
  
### Plot UMAP of human and mouse IT types here
  
#### Human
  
First, select variable features based on beta score (this replaces the default Seurat method and is also independent of the Patch-seq cells).  
  
```{r variable gene selection}
nGenes     <- 2000
cl         <- setNames(annoMTG$cluster_label,colnames(datMTG))
propExpr   <- get_cl_prop(datMTG,cl)
betaScore  <- getBetaScore(propExpr)
betaOutH   <- data.frame(Gene=rownames(datMTG),BetaScore=betaScore)
betaOutH   <- betaOutH[order(-betaScore),]
varFeatures<- betaOutH$Gene[1:nGenes]
```
  
  
Second build the human Seurat object and run the preparations.   
  
```{r human seurat}
spl <- function(a) as.character(lapply(a,function(x) strsplit(x," ")[[1]][4]))
human_metadata <- data.frame(celltype = spl(annoMTG$cluster_label),
                             clusterFrem = spl(annoMTG$frem_layer),
                             layer = annoMTG$layer_label,
                             subclusterFrem = annoMTG$frem_subtype)
rownames(human_metadata) <- colnames(datMTG)
human <- CreateSeuratObject(datMTG, meta.data = human_metadata)
human <- NormalizeData(human, verbose = FALSE)
VariableFeatures(human) <- varFeatures
human <- ScaleData(object = human, verbose = FALSE)
human <- RunPCA(object = human, npcs = 30, verbose = FALSE)
human <- RunUMAP(object = human, reduction = "pca", dims = 1:30, verbose = FALSE)#, umap.method = 'umap-learn')#, metric = 'correlation')
``` 
  
  
Third, plot the human umap.  **Important note: UMAPs may look slightly different with different R configurations due to how the umap functionality is implemented in Seurat.  These minor umap differences do not impact manuscript conclusions.**  
  
```{r plot human seurat, fig.height=5,fig.width=15}
xl <- range(FetchData(human,"UMAP_1"))
yl <- range(FetchData(human,"UMAP_2"))
cols <- unique(annoMTG$cluster_color)[match(sort(unique(human_metadata$celltype)),unique(human_metadata$celltype))]
f1 <- colnames(human)[annoMTG$frem_subtype!="other"]
pa1 <- DimPlot(object = human, group.by = "subclusterFrem", reduction = "umap", pt.size = 1, 
              label=TRUE, label.size = 8, cells=f1) + NoLegend() + xlim(xl) + ylim(yl)
pa2 <- DimPlot(object = human, group.by = "celltype", reduction = "umap", pt.size = 1, 
              label=TRUE, cols=cols) + NoLegend() + xlim(xl) + ylim(yl)
f3 <- colnames(human)[annoMTG$cluster_label=="Exc L2-3 LINC00507 FREM3"]
pa3 <- DimPlot(object = human, group.by = "layer",  reduction = "umap", pt.size = 1, 
              label=TRUE, label.size = 8, cells=f3) + NoLegend() + xlim(xl) + ylim(yl)
plot_grid(pa1, pa2, pa3, ncol = 3)
p_human <- pa2 # For saving coordinates later

## Repeat with labels set to FALSE
xl <- range(FetchData(human,"UMAP_1"))
yl <- range(FetchData(human,"UMAP_2"))
cols <- unique(annoMTG$cluster_color)[match(sort(unique(human_metadata$celltype)),unique(human_metadata$celltype))]
f1 <- colnames(human)[annoMTG$frem_subtype!="other"]
p1 <- DimPlot(object = human, group.by = "subclusterFrem", reduction = "umap", pt.size = 1, 
              label=FALSE, label.size = 8, cells=f1) + NoLegend() + xlim(xl) + ylim(yl)
p2 <- DimPlot(object = human, group.by = "celltype", reduction = "umap", pt.size = 1, 
              label=FALSE, cols=cols) + NoLegend() + xlim(xl) + ylim(yl)
f3 <- colnames(human)[annoMTG$cluster_label=="Exc L2-3 LINC00507 FREM3"]
p3 <- DimPlot(object = human, group.by = "layer",  reduction = "umap", pt.size = 1, 
              label=FALSE, label.size = 8, cells=f3) + NoLegend() + xlim(xl) + ylim(yl)
plot_grid(p1, p2, p3, ncol = 3)
ggsave("UMAP_human.pdf",height=6,width=15)

## Output in a different format
p <- list()
p[[1]] <- list(); p[[1]][[1]] <- pa1
p[[2]] <- list(); p[[2]][[1]] <- pa2
p[[3]] <- list(); p[[3]][[1]] <- pa3
p[[4]] <- list(); p[[4]][[1]] <- p1
p[[5]] <- list(); p[[5]][[1]] <- p2
p[[6]] <- list(); p[[6]][[1]] <- p3
pdf("umap_human_one_per_page.pdf", onefile = TRUE, height=4.5, width=5)
for (i in seq(length(p))) {
  do.call("grid.arrange", p[[i]])  
}
dev.off()
```  
  
Use genesorteR to identify genes differentially expressed between FREM3, COL22A1, and CARM1P1 for DEX analysis.  
  
```{r}
#filter cells
cl3       <- c("Exc L2-3 LINC00507 FREM3", "Exc L3-4 RORB CARM1P1", "Exc L3-5 RORB COL22A1")
kp3       <- is.element(annoMTG$cluster_label,cl3)&
  ((is.element(annoMTG$frem_subtype,c("f73"))&is.element(annoMTG$layer_label,c("L3","L4")))|  #"f79"
     (annoMTG$cluster_label!="Exc L2-3 LINC00507 FREM3"))
propExpr3 <- get_cl_prop(datMTG[,kp3],cl[kp3])


#very roughly filter genes
betaScore3 <- getBetaScore(propExpr3[rowSums(propExpr3[,cl3])>0.3,cl3])
kpGenes3   <- names(sort(-betaScore3))[1:10000]
write(sort(kpGenes3[1:300]),"human_markers_beta300.txt")

#get specificity scores for each cell type
sg = sortGenes(datMTG[kpGenes3,kp3], annoMTG$cluster_label[kp3])

#define a small set of markers
mm3 = getMarkers(sg, quant = 0.7)

#cluster genes and make a heatmap
pp = plotMarkerHeat(sg$inputMat, sg$inputClass, mm3$markers, clusterGenes=TRUE, outs = TRUE)
pdf("dex_heatmap_genesorteR.pdf",height = 15,width=8)
pp$pheat
dev.off()

pp = plotMarkerHeat(sg$inputMat, sg$inputClass, kpGenes3[1:300], clusterGenes=TRUE, outs = TRUE)
pdf("dex_heatmap_beta300.pdf",height = 19,width=8)
pp$pheat
dev.off()

write(sort(mm3$markers),"human_markers_genesorteR.txt")
```
  
  
  
Now use genesorteR to identify genes differentially expressed between deep and superficial L2/3.  
  
```{r}
#filter cells
cl2       <- c("Exc L2-3 LINC00507 FREM3", "Exc L2-4 LINC00507 GLP2R", "Exc L2 LAMP5 LTK")
kp2       <- is.element(annoMTG$cluster_label,cl2)&
  ((is.element(annoMTG$layer_label,c("L2")))|(annoMTG$cluster_label!="Exc L2-3 LINC00507 FREM3"))
kp        <- kp2|kp3

depth     <- setNames(c("deep","superficial")[kp2+1],colnames(datMTG))
propExpr2 <- get_cl_prop(datMTG[,kp],depth[kp])


#very roughly filter genes
betaScore2 <- getBetaScore(propExpr2)
kpGenes2   <- names(sort(-betaScore2))[1:10000]
write(sort(kpGenes2[1:300]),"human_markers_L23_beta300.txt")

#get specificity scores for each cell type
sg = sortGenes(datMTG[kpGenes2,kp], depth[kp])

#define a small set of markers
mm = getMarkers(sg, quant = 0.7)

#cluster genes and make a heatmap
pp = plotMarkerHeat(sg$inputMat, sg$inputClass, mm$markers, clusterGenes=TRUE, outs = TRUE)
pdf("dex_heatmap_L23_genesorteR.pdf",height = 15,width=8)
pp$pheat
dev.off()

pp = plotMarkerHeat(sg$inputMat, sg$inputClass, kpGenes3[1:300], clusterGenes=TRUE, outs = TRUE)
pdf("dex_heatmap_L23_beta300.pdf",height = 19,width=8)
pp$pheat
dev.off()

write(sort(mm$markers),"human_markers_L23_genesorteR.txt")
```
  
  
Plot some canonical genes in this space for annotation.  In particular, we want to plot all genes that are selectively expressed for various deep layer 3 types.  We also plot FOXP2, which is one of the genes distinguising superficual L2/3 from deep L3.  
  
```{r genes plot2, fig.height=7,fig.width=14, message=FALSE, warning=FALSE}
features = c("RORB","LAMP5",'CARM1P1',"COL22A1","FREM3","LTK","GLP2R","CUX2")
FeaturePlot(object = human, reduction = "umap", pt.size = 0.5, 
                        features = features, ncol=4, slot="counts") + 
    ylim(xl) + ylim(yl)

features = c("NEFH",sort(mm3$markers))  # Also including NEFH for inclusion in figure

p<-list()
for (i in 1:length(features)) {
  p[[i]] <- list()
  p[[i]][[1]] <- FeaturePlot(object = human, reduction = "umap", pt.size = 0.5, 
                        features = features[i], ncol=1, slot="counts") + 
    ylim(xl) + ylim(yl)
}

pdf("genePlots_deepTypes_L3genes.pdf", onefile = TRUE, height=4.5, width=5)
for (i in seq(length(p))) {
  do.call("grid.arrange", p[[i]])  
}
dev.off()

pdf("genePlots_deepTypes_L3genes_Fig4.pdf", onefile = TRUE, height=4.5, width=5)
fig4genes <- which(is.element(features,c("COBLL1","KCNK2","NEFH","NTN4","PHLDB2","QRFPR","SEMA5A","SV2C")))
for (i in fig4genes) {
  do.call("grid.arrange", p[[i]])  
}
dev.off()
```
  
  
#### Mouse VISp
  
First, select variable features based on beta score (this replaces the default Seurat method and is also independent of the Patch-seq cells).  
  
```{r mouse VISp variable gene selection}
nGenes     <- 2000
cl         <- setNames(annoVISp$cluster_label,colnames(datVISp))
propExpr   <- get_cl_prop(datVISp,cl)
betaScore  <- getBetaScore(propExpr)
betaOutM   <- data.frame(Gene=rownames(datVISp),BetaScore=betaScore)
betaOutM   <- betaOutM[order(-betaScore),]
varFeatures<- betaOutM$Gene[1:nGenes]
```
  
  
Second build the mouse VISp Seurat object and run the preparations.  
  
```{r mouse VISp seurat}
spl2 <- function(a) as.character(lapply(a,function(x) {a=strsplit(x," ")[[1]]; paste(a[c(1,length(a))],collapse = " ")}))
mouse_metadata <- data.frame(celltype = spl2(annoVISp$cluster_label),
                             clusterFrem = annoVISp$cluster_label)
rownames(mouse_metadata) <- colnames(datVISp)
mouseV <- CreateSeuratObject(datVISp, meta.data = mouse_metadata)
mouseV <- NormalizeData(mouseV, verbose = FALSE)
#mouseV <- FindVariableFeatures(mouseV, selection.method = "vst", nfeatures = 2000, verbose = FALSE) # 750
VariableFeatures(mouseV) <- varFeatures
mouseV <- ScaleData(object = mouseV, verbose = FALSE)
mouseV <- RunPCA(object = mouseV, npcs = 30, verbose = FALSE)
mouseV <- RunUMAP(object = mouseV, reduction = "pca", dims = 1:30, verbose = FALSE)
``` 


```{r plot mouse VISp seurat, fig.height=5,fig.width=15}
colsM <- unique(annoVISp$cluster_color)[match(sort(unique(mouse_metadata$celltype)),unique(mouse_metadata$celltype))]
p1 <- DimPlot(object = mouseV, group.by = "celltype", reduction = "umap", pt.size = 1, 
              label=TRUE, cols=colsM) + NoLegend() + xlim(xl) + ylim(yl)
p2 <- DimPlot(object = mouseV, group.by = "celltype", reduction = "umap", pt.size = 1, 
              label=FALSE, cols=colsM) + NoLegend() + xlim(xl) + ylim(yl)
p_mouseV <- p2 # For saving coordinates later
plot_grid(p1, p2, ncol = 3)
ggsave("UMAP_mouse_VISp.pdf",height=6,width=15)
```  
  
  
#### Mouse ALM
  
First, select variable features based on beta score (this replaces the default Seurat method and is also independent of the Patch-seq cells).  
  
```{r mouse ALM variable gene selection}
nGenes     <- 2000
cl         <- setNames(annoALM$cluster_label,colnames(datALM))
propExpr   <- get_cl_prop(datALM,cl)
betaScore  <- getBetaScore(propExpr)
betaOutA   <- data.frame(Gene=rownames(datALM),BetaScore=betaScore)
betaOutA   <- betaOutM[order(-betaScore),]
varFeatures<- betaOutA$Gene[1:nGenes]
```
  
  
Second build the mouse ALM Seurat object and run the preparations.  
  
```{r mouse ALM seurat}
spl2 <- function(a) as.character(lapply(a,function(x) {a=strsplit(x," ")[[1]]; paste(a[c(1,length(a))],collapse = " ")}))
mouse_metadata <- data.frame(celltype = spl2(annoALM$cluster_label),
                             clusterFrem = annoALM$cluster_label)
rownames(mouse_metadata) <- colnames(datALM)
mouseA <- CreateSeuratObject(datALM, meta.data = mouse_metadata)
mouseA <- NormalizeData(mouseA, verbose = FALSE)
#mouseA <- FindVariableFeatures(mouseA, selection.method = "vst", nfeatures = 2000, verbose = FALSE) # 750
VariableFeatures(mouseA) <- varFeatures
mouseA <- ScaleData(object = mouseA, verbose = FALSE)
mouseA <- RunPCA(object = mouseA, npcs = 30, verbose = FALSE)
mouseA <- RunUMAP(object = mouseA, reduction = "pca", dims = 1:30, verbose = FALSE)
``` 


```{r plot mouse ALM seurat, fig.height=5,fig.width=15}
colsM <- unique(annoALM$cluster_color)[match(sort(unique(mouse_metadata$celltype)),unique(mouse_metadata$celltype))]
p1 <- DimPlot(object = mouseA, group.by = "celltype", reduction = "umap", pt.size = 1, 
              label=TRUE, cols=colsM) + NoLegend() + xlim(xl) + ylim(yl)
p2 <- DimPlot(object = mouseA, group.by = "celltype", reduction = "umap", pt.size = 1, 
              label=FALSE, cols=colsM) + NoLegend() + xlim(xl) + ylim(yl)
p_mouseA <- p2 # For saving coordinates later
plot_grid(p1, p2, ncol = 3)
ggsave("UMAP_mouse_ALM.pdf",height=6,width=15)
```  
  
  
Save UMAP coordinates for human and mouse.  These will be used for mapping Patch-seq data in later scripts.  
  
```{r save umap}
umap_human  <- p_human$data[,c("UMAP_1","UMAP_2")]
umap_mouseV <- p_mouseV$data[,c("UMAP_1","UMAP_2")]
umap_mouseA <- p_mouseA$data[,c("UMAP_1","UMAP_2")]
save(umap_human,umap_mouseV,umap_mouseA,file="umap_coordinates.RData")
```
  
  
### Assess cluster heterogeneity in mouse VISp/ALM and human MTG  
  
To assess cluster heterogeneity, we will compare the variance explained from the first PC in real data against the variance epxlained from the first PC in random shufflings of the data.  Clusters with high heterogeneity should also have at least one signficant PC relative to the expected PC value from chance.  We will downsample to the same number of cells per cluster and use the same number of variable genes calculated the same way in mouse and human for a fair comparison.  We are also excluding outlier samples from the PC calculation to insure that the results are not driven by a small number of outlier or low quality cells, and are excluding sex and mitochondrial genes.  
  
First some housekeeping things (functions, common and non-excluded genes).  
  
```{r define required functions}
## Select the genes of common mouse/human genes that are not sex or mitochondrial genes for later
data(mito_genes)  
data(sex_genes)  
excludeGenes <- sort(unique(c(sex_genes,mito_genes)))  
kpGenes <- intersect(rownames(datVISp),rownames(datMTG))
kpGenes <- sort(setdiff(kpGenes,excludeGenes))

## Required functions
findFromGroupsVector <- function(x,y,...) { 
 x = as.numeric(x)
 x = cbind(x,x)
 f = findFromGroups(x,y,...)
 return(colMeans(f))
}

# This function calculates the top PCs after removing any outlier cells 
# (to account for potential low-quality or misassigned cells)
prcompAfterOutlierRemoval <- function(tnorm.dat,gnUse=1:dim(tnorm.dat)[2],kpCl=1:dim(tnorm.dat)[2],zThresh){
  pcaUse <- prcomp(tnorm.dat[kpCl,gnUse])
  go = TRUE
  # Check for outliers and recalculate PCs until no outliers remain.
  while(go&length(kpCl>10)){
    z1 = pcaUse$x[,1];  z1 = (z1-mean(z1))/sd(z1)
    z2 = pcaUse$x[,2];  z2 = (z2-mean(z2))/sd(z2)
	# Check for outliers
    kpTmpS = pmax(abs(z1),abs(z2))<zThresh
	if (sum(!kpTmpS)==0){
	  go = FALSE
	} else {
      kpCl[kpCl][!kpTmpS] = FALSE
      pcaUse <- prcomp(tnorm.dat[kpCl,gnUse])
    } 
  }
  return(pcaUse)
}
```
  
  
Next, find the actual and permuted PCs in human.  
  
```{r human scramble PCs}
set.seed(1)
nCells  <- 70     # Decreasing from 80 to 70 to match the minimum size of ALM clusters (L2/3 IT ALM Ptrf)
nGenes  <- nCells
rCount  <- 10 # 2
pCount  <- 100
zThresh <- 3
pcMax   <- 10
p1h <- p1hr <- list()
clustsH <- sort(unique(annoMTG$cluster_label))[c(3,2,1,4,5)] # Order superficial to deep
for(cls in clustsH){
  print(cls)
  for (i in 1:rCount){
    if(i==1) {
      p1h[[cls]]  <- NULL
      p1hr[[cls]] <- NULL
    }
    set.seed(i)
    kpCell <- sample(colnames(datMTG)[annoMTG$cluster_label==cls],nCells)
    nbt    <- CreateSeuratObject(datMTG[kpGenes,kpCell])
    nbt    <- FindVariableFeatures(nbt, selection.method = "vst", nfeatures = nGenes, mean.cutoff = c(1,Inf), verbose = FALSE)
    datTmp <- as.matrix(FetchData(nbt,vars = VariableFeatures(nbt), slot="counts"))
    pcaUse <- prcompAfterOutlierRemoval(datTmp,zThresh=zThresh)
    varExp <- ((10000*(pcaUse$sdev)^2 / sum(pcaUse$sdev^2))/100)[1:pcMax]
    p1h[[cls]] <- rbind(p1h[[cls]],varExp)
    for (j in 1:pCount){
      set.seed(i*pCount+j)
      datRand <- t(apply(datTmp,1,function(x) return(x[sample(1:length(x),length(x))])))
      colnames(datRand) <- colnames(datTmp)
	    pcaRand <- prcompAfterOutlierRemoval(datRand,zThresh=zThresh)
	    varExpR <- ((10000*(pcaRand$sdev)^2 / sum(pcaRand$sdev^2))/100)[1:pcMax]
	    p1hr[[cls]] <- rbind(p1hr[[cls]],varExp/varExpR)
    }
  }
}
```
  
  
Repeat, finding the actual and permuted PCs in mouse VISp.  
  
```{r mouse VISp scramble PCs}
clustsM <- sort(unique(annoVISp$cluster_label))[c(1,3,2)] # Order superficial to deep
for(cls in clustsM){
  print(cls)
  for (i in 1:rCount){
    if(i==1) {
      p1h[[cls]]  <- NULL
      p1hr[[cls]] <- NULL
    }
    set.seed(i)
    kpCell <- sample(colnames(datVISp)[annoVISp$cluster_label==cls],nCells)
    nbt    <- CreateSeuratObject(datVISp[kpGenes,kpCell])
    nbt    <- FindVariableFeatures(nbt, selection.method = "vst", nfeatures = nGenes, mean.cutoff = c(1,Inf), verbose = FALSE)
    datTmp <- as.matrix(FetchData(nbt,vars = VariableFeatures(nbt), slot="counts"))
    pcaUse <- prcompAfterOutlierRemoval(datTmp,zThresh=zThresh)
    varExp <- ((10000*(pcaUse$sdev)^2 / sum(pcaUse$sdev^2))/100)[1:pcMax]
    p1h[[cls]] <- rbind(p1h[[cls]],varExp)
    for (j in 1:pCount){
      set.seed(i*pCount+j)
      datRand <- t(apply(datTmp,1,function(x) return(x[sample(1:length(x),length(x))])))
      colnames(datRand) <- colnames(datTmp)
	    pcaRand <- prcompAfterOutlierRemoval(datRand,zThresh=zThresh)
	    varExpR <- ((10000*(pcaRand$sdev)^2 / sum(pcaRand$sdev^2))/100)[1:pcMax]
	    p1hr[[cls]] <- rbind(p1hr[[cls]],varExp/varExpR)
    }
  }
}
```  
  
Repeat, finding the actual and permuted PCs in mouse ALM.  
  
```{r mouse ALM scramble PCs}
clustsA <- sort(unique(annoALM$cluster_label))[c(1,3,2)] # No specific order
for(cls in clustsA){
  print(cls)
  for (i in 1:rCount){
    if(i==1) {
      p1h[[cls]]  <- NULL
      p1hr[[cls]] <- NULL
    }
    set.seed(i)
    kpCell <- sample(colnames(datALM)[annoALM$cluster_label==cls],nCells)
    nbt    <- CreateSeuratObject(datALM[kpGenes,kpCell])
    nbt    <- FindVariableFeatures(nbt, selection.method = "vst", nfeatures = nGenes, mean.cutoff = c(1,Inf), verbose = FALSE)
    datTmp <- as.matrix(FetchData(nbt,vars = VariableFeatures(nbt), slot="counts"))
    pcaUse <- prcompAfterOutlierRemoval(datTmp,zThresh=zThresh)
    varExp <- ((10000*(pcaUse$sdev)^2 / sum(pcaUse$sdev^2))/100)[1:pcMax]
    p1h[[cls]] <- rbind(p1h[[cls]],varExp)
    for (j in 1:pCount){
      set.seed(i*pCount+j)
      datRand <- t(apply(datTmp,1,function(x) return(x[sample(1:length(x),length(x))])))
      colnames(datRand) <- colnames(datTmp)
	    pcaRand <- prcompAfterOutlierRemoval(datRand,zThresh=zThresh)
	    varExpR <- ((10000*(pcaRand$sdev)^2 / sum(pcaRand$sdev^2))/100)[1:pcMax]
	    p1hr[[cls]] <- rbind(p1hr[[cls]],varExp/varExpR)
    }
  }
}
```  
  
  
  
Now plot the results, comparing variation across all 11 types.  
  
```{r plot heterogeneity results, fig.height=7,fig.width=9}
meanDiff <- sdDiff <- NULL
for (cls in names(p1hr)){
  meanDiff <- rbind(meanDiff,colMeans(p1hr[[cls]]))
  sdDiff   <- rbind(sdDiff, apply(p1hr[[cls]],2,sd))
}
rownames(meanDiff) <- rownames(sdDiff) <- names(p1hr)
colnames(meanDiff) <- colnames(sdDiff) <- paste0("PC",1:pcMax)

# Data
data <- data.frame(
  name=factor(c(spl(clustsH),spl2(clustsM),spl2(clustsA)),levels=c(spl(clustsH),spl2(clustsM),spl2(clustsA))),
  value=meanDiff[,1],
  sd=sdDiff[,1]
)

a = setNames(c(annoVISp$cluster_color,annoALM$cluster_color,annoMTG$cluster_color),
             c(annoVISp$cluster_label,annoALM$cluster_label,annoMTG$cluster_label))
data$cols <- as.character(a[rownames(data)])

# line + dot
ggplot(data) +
  geom_bar( aes(x=name, y=value), stat="identity", fill=data$cols, alpha=1)+#0.5) +
  geom_errorbar( aes(x=name, ymin=value-sd, ymax=value+sd), width=0.4, colour="black", alpha=0.9, size=1.3) +
  ylab("Actual/permuted variance explained by PC1") + xlab("") + theme_bw() +
  theme(axis.text.x = element_text(angle = 90), text = element_text(size=16))
ggsave("heterogeneity.pdf",height=8,width=6)
```
  
  
## Cluster discreteness (based on number of pairwise DEX genes)
  
Using `scrattch.hi`, calculating the number of DEX genes between each pair of clusters, subsampling to 80 cells per cluster and repeating averaging across 10 iterations again.  
  
```{r dex genes using hicat}
de.param   <- de_param(low.th = 1, padj.th = 0.01, lfc.th = 1,    # Recommended parameters for SMART-seq data
                       q1.th = 0.5, q.diff.th = 0.7, de.score.th = 150)
humanCount <- mouseCountV <- mouseCountA <- 0
numIters   <- 10

for (n in 1:numIters){
  ## Human
  cl <- setNames(make.names(annoMTG$cluster_label),annoMTG$sample_id)
  nm <- sort(unique(cl))
  kpSamp <- subsampleCells(cl,subSamp=nCells,seed=n)
  human_de <- de_score(datMTG[kpGenes, kpSamp], cl[kpSamp], de.param)
  human_count <- matrix(0, ncol = length(nm), nrow = length(nm))
  rownames(human_count) <- colnames(human_count) <- nm
  for (i in 1:(length(nm)-1)) for (j in (i+1):length(nm)){
    human_count[i,j] = human_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$up.num
    human_count[j,i] = human_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$down.num
  }
  humanCount = human_count + humanCount
  
  ## Mouse VISp
  cl <- setNames(make.names(annoVISp$cluster_label),annoVISp$sample_id)
  nm <- sort(unique(cl))
  kpSamp <- subsampleCells(cl,subSamp=nCells,seed=n)
  mouse_de <- de_score(datVISp[kpGenes, kpSamp], cl[kpSamp], de.param)
  mouse_count <- matrix(0, ncol = length(nm), nrow = length(nm))
  rownames(mouse_count) <- colnames(mouse_count) <- nm
  for (i in 1:(length(nm)-1)) for (j in (i+1):length(nm)){
    mouse_count[i,j] = mouse_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$up.num
    mouse_count[j,i] = mouse_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$down.num
  }
  mouseCountV = mouse_count + mouseCountV
  
  ## Mouse ALM
  cl <- setNames(make.names(annoALM$cluster_label),annoALM$sample_id)
  nm <- sort(unique(cl))
  kpSamp <- subsampleCells(cl,subSamp=nCells,seed=n)
  mouse_de <- de_score(datALM[kpGenes, kpSamp], cl[kpSamp], de.param)
  mouse_count <- matrix(0, ncol = length(nm), nrow = length(nm))
  rownames(mouse_count) <- colnames(mouse_count) <- nm
  for (i in 1:(length(nm)-1)) for (j in (i+1):length(nm)){
    mouse_count[i,j] = mouse_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$up.num
    mouse_count[j,i] = mouse_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$down.num
  }
  mouseCountA = mouse_count + mouseCountA
}
mouseCountV = (mouseCountV+t(mouseCountV))/numIters
mouseCountA = (mouseCountA+t(mouseCountA))/numIters
humanCount  = (humanCount+t(humanCount))/numIters
print(paste("Human     :",sum(humanCount[1:3,1:3])/6))
print(paste("Mouse VISp:",sum(mouseCountV[1:3,1:3])/6))
print(paste("Mouse ALM :",sum(mouseCountA[1:3,1:3])/6))
```
  
  
Comparable DEX genes in matched L2/3 types in mouse and human (maybe fewer in human) after matching for number of cells.  Plot these results.  
  
```{r plot dex genes}
out <- t(cbind(humanCount[1:3,],mouseCountV,mouseCountA))
colnames(out) <- 1:3
out[out==0]=NA
out <- as.data.frame(out)
out$name <- rownames(out) <- data$name

barplot(apply(out[,1:3],1,mean,na.rm=TRUE),las=2,col=data$cols, ylab="# DEX genes wrt matched human/mouse types",border="white")

pdf("discreteness.pdf",height=8,width=4)
barplot(apply(out[,1:3],1,mean,na.rm=TRUE),las=2,col=data$cols, ylab="# DEX genes wrt matched human/mouse types",border="white")
dev.off()
```
  
  
## Expression of mFISH markers in FACS
  
Finally, let's plot the expected cell type specificity of marker genes that are used in mFISH experiments to assess laminar localization of cell types in situ.  Note that an older interacitve version of the scripts were used to make the version of the plot in Extended Data Figure 3d, but that the overall pattern (and thus conclusions based on this plot) are entirely unchanged.   
  
```{r plot heatmap, fig.height=6,fig.width=4}
mark    <- c("LAMP5","LTK","CUX2","GLP2R","RORB","FREM3","CARTPT","COL22A1")
data.df <- cbind(sample_name = colnames(datMTG), as.data.frame(2^t(datMTG[mark,])-1))
anno    <- data.frame(sample_name=as.character(annoMTG$sample_id),
                      cluster_label=as.character(annoMTG$cluster_label),
                      cluster_color=as.character(annoMTG$cluster_color))
anno$cluster_id <- as.numeric(factor(anno$cluster_label,levels=clustsH))
scrattch.vis::group_heatmap_plot(data.df, anno, genes=mark, grouping ="cluster", stat="median",
                                 log_scale = TRUE, max_width = 30, label_height = 35)
```

  
Output session information.  
  
```{r sessionInfo}
sessionInfo()
```      
  