---
title: "Cross-species analysis"
author: "Jeremy Miller"
date: "March 27, 2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This script performs some statistical analyses and makes some UMAP plots showing glutamatergic neurons in supragranular layers of human MTG and mouse VISp.  One purpose is to quantify to what extent the cell type diversity, variation, and specialization of human types extends beyond that of mouse.  Note that this script is only considering dissociated cells/nuclei (called "FACS" here) and includes code for generating panels of Figure 1.   
If needed set the working directory:  (e.g., `setwd("C:/Users/jeremym/Desktop/patchseq_human_L23")`). 
  
### Work space preparation
  
First, load all the libraries (required for all code documents) and set some options.  
  
```{r load libraries, warning=FALSE}
# Load all packages
suppressPackageStartupMessages({
  library(genesorteR)
  library(feather)
  library(dplyr)
  library(gplots)
  library(Seurat)
  library(ggplot2)
  library(cowplot)
  library(scrattch.hicat)
  library(future)
  library(ggbeeswarm)
  library(mfishtools) # getBetaScore
  library(pheatmap)
  library(VENcelltypes) # mito_genes and sex_genes
  library(WGCNA)
  library(Matrix)
  library(gridExtra)
})
options(stringsAsFactors=FALSE)
options(future.globals.maxSize = 4000 * 1024^2)
```
  
  
Next, load the all the relevant data, which were calculated in a previous script.  
  
```{r load data}
load("input_data_sets.RData")
```
  
  
### Plot UMAP of human and mouse IT types here
  
#### Human
  
First, select variable features based on beta score (this replaces the default Seurat method and is also independent of the Patch-seq cells).  
  
```{r variable gene selection}
nGenes     <- 2000
cl         <- setNames(annoMTG$cluster_label,colnames(datMTG))
propExpr   <- get_cl_prop(datMTG,cl)
betaScore  <- getBetaScore(propExpr)
betaOutH   <- data.frame(Gene=rownames(datMTG),BetaScore=betaScore)
betaOutH   <- betaOutH[order(-betaScore),]
varFeatures<- betaOutH$Gene[1:nGenes]
```
  
  
Second build the human Seurat object and run the preparations.   
  
```{r human seurat}
spl <- function(a) as.character(lapply(a,function(x) strsplit(x," ")[[1]][4]))
human_metadata <- data.frame(celltype = spl(annoMTG$cluster_label),
                             clusterFrem = spl(annoMTG$frem_layer),
                             layer = annoMTG$layer_label,
                             subclusterFrem = annoMTG$frem_subtype)
rownames(human_metadata) <- colnames(datMTG)
human <- CreateSeuratObject(datMTG, meta.data = human_metadata)
human <- NormalizeData(human, verbose = FALSE)
VariableFeatures(human) <- varFeatures
human <- ScaleData(object = human, verbose = FALSE)
human <- RunPCA(object = human, npcs = 30, verbose = FALSE)
human <- RunUMAP(object = human, reduction = "pca", dims = 1:30, verbose = FALSE)
``` 
  
  
Third, plot the human umap.  
  
```{r plot human seurat, fig.height=5,fig.width=15}
xl <- range(FetchData(human,"UMAP_1"))
yl <- range(FetchData(human,"UMAP_2"))
cols <- unique(annoMTG$cluster_color)[match(sort(unique(human_metadata$celltype)),unique(human_metadata$celltype))]
f1 <- colnames(human)[annoMTG$frem_subtype!="other"]
p1 <- DimPlot(object = human, group.by = "subclusterFrem", reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=TRUE, label.size = 8, cells=f1) + NoLegend() + xlim(xl) + ylim(yl)
p2 <- DimPlot(object = human, group.by = "celltype", reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=TRUE, cols=cols) + NoLegend() + xlim(xl) + ylim(yl)
f3 <- colnames(human)[annoMTG$cluster_label=="Exc L2-3 LINC00507 FREM3"]
p3 <- DimPlot(object = human, group.by = "layer",  reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=TRUE, label.size = 8, cells=f3) + NoLegend() + xlim(xl) + ylim(yl)
plot_grid(p1, p2, p3, ncol = 3)

## Repeat with labels set to FALSE
xl <- range(FetchData(human,"UMAP_1"))
yl <- range(FetchData(human,"UMAP_2"))
cols <- unique(annoMTG$cluster_color)[match(sort(unique(human_metadata$celltype)),unique(human_metadata$celltype))]
f1 <- colnames(human)[annoMTG$frem_subtype!="other"]
p1 <- DimPlot(object = human, group.by = "subclusterFrem", reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=FALSE, label.size = 8, cells=f1) + NoLegend() + xlim(xl) + ylim(yl)
p2 <- DimPlot(object = human, group.by = "celltype", reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=FALSE, cols=cols) + NoLegend() + xlim(xl) + ylim(yl)
f3 <- colnames(human)[annoMTG$cluster_label=="Exc L2-3 LINC00507 FREM3"]
p3 <- DimPlot(object = human, group.by = "layer",  reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=FALSE, label.size = 8, cells=f3) + NoLegend() + xlim(xl) + ylim(yl)
plot_grid(p1, p2, p3, ncol = 3)
ggsave("UMAP_human.pdf",height=6,width=15)
```  
  
Use genesorteR to identify genes differentially expressed between FREM3, COL22A1, and CARM1P1 for DEX analysis.  
  
```{r}
#filter cells
cl3       <- c("Exc L2-3 LINC00507 FREM3", "Exc L3-4 RORB CARM1P1", "Exc L3-5 RORB COL22A1")
kp3       <- is.element(annoMTG$cluster_label,cl3)&
  ((is.element(annoMTG$frem_subtype,c("f73"))&is.element(annoMTG$layer_label,c("L3","L4")))|  #"f79"
     (annoMTG$cluster_label!="Exc L2-3 LINC00507 FREM3"))
propExpr3 <- get_cl_prop(datMTG[,kp3],cl[kp3])


#very roughly filter genes
betaScore3 <- getBetaScore(propExpr[rowSums(propExpr[,cl3])>0.3,cl3])
kpGenes3   <- names(sort(-betaScore3))[1:10000]
write(sort(kpGenes3[1:300]),"human_markers_beta300.txt")

#get specificity scores for each cell type
sg = sortGenes(datMTG[kpGenes3,kp3], annoMTG$cluster_label[kp3])

#define a small set of markers
mm = getMarkers(sg, quant = 0.7)

#cluster genes and make a heatmap
pp = plotMarkerHeat(sg$inputMat, sg$inputClass, mm$markers, clusterGenes=TRUE, outs = TRUE)
pdf("dex_heatmap_genesorteR.pdf",height = 15,width=8)
pp$pheat
dev.off()

pp = plotMarkerHeat(sg$inputMat, sg$inputClass, kpGenes3[1:300], clusterGenes=TRUE, outs = TRUE)
pdf("dex_heatmap_beta300.pdf",height = 19,width=8)
pp$pheat
dev.off()

write(sort(mm$markers),"human_markers_genesorteR.txt")
```
  
  
Plot some canonical genes in this space for annotation.  In particular, we want to plot all genes that are selectively expressed for various deep layer 3 types.  The file read in below is a superset of the genes plotted in Figure 6f and includes all chosen by genesorteR based on FACS (above) and Patch-seq (in a later script).  
  
```{r genes plot2, fig.height=7,fig.width=14, message=FALSE, warning=FALSE}
features = c("RORB","LAMP5",'CARM1P1',"COL22A1","FREM3","LTK","GLP2R","CUX2")
FeaturePlot(object = human, reduction = "umap", pt.size = 0.5, 
                        features = features, ncol=4, slot="counts") + 
    ylim(xl) + ylim(yl)

features = scan("data/human_markers_beta300_FACSandPatch.txt",what="character",sep="\n")

p<-list()
for (i in 1:length(features)) {
  p[[i]] <- list()
  p[[i]][[1]] <- FeaturePlot(object = human, reduction = "umap", pt.size = 0.5, 
                        features = features[i], ncol=1, slot="counts") + 
    ylim(xl) + ylim(yl)
}

pdf("genePlots_deepTypes_beta300.pdf", onefile = TRUE, height=4.5, width=5)
for (i in seq(length(p))) {
  do.call("grid.arrange", p[[i]])  
}
dev.off()
```
  
  
#### Mouse
  
First, select variable features based on beta score (this replaces the default Seurat method and is also independent of the Patch-seq cells).  
  
```{r mouse variable gene selection}
nGenes     <- 2000
cl         <- setNames(annoVISp$cluster_label,colnames(datVISp))
propExpr   <- get_cl_prop(datVISp,cl)
betaScore  <- getBetaScore(propExpr)
betaOutM   <- data.frame(Gene=rownames(datVISp),BetaScore=betaScore)
betaOutM   <- betaOutM[order(-betaScore),]
varFeatures<- betaOutM$Gene[1:nGenes]
```
  
  
Second build the mouse Seurat object and run the preparations.  
  
```{r mouse seurat}
spl2 <- function(a) as.character(lapply(a,function(x) {a=strsplit(x," ")[[1]]; paste(a[c(1,length(a))],collapse = " ")}))
mouse_metadata <- data.frame(celltype = spl2(annoVISp$cluster_label),
                             clusterFrem = annoVISp$cluster_label)
rownames(mouse_metadata) <- colnames(datVISp)
mouse <- CreateSeuratObject(datVISp, meta.data = mouse_metadata)
mouse <- NormalizeData(mouse, verbose = FALSE)
#mouse <- FindVariableFeatures(mouse, selection.method = "vst", nfeatures = 2000, verbose = FALSE) # 750
VariableFeatures(mouse) <- varFeatures
mouse <- ScaleData(object = mouse, verbose = FALSE)
mouse <- RunPCA(object = mouse, npcs = 30, verbose = FALSE)
mouse <- RunUMAP(object = mouse, reduction = "pca", dims = 1:30, verbose = FALSE)
``` 


```{r plot mouse seurat, fig.height=5,fig.width=15}
colsM <- unique(annoVISp$cluster_color)[match(sort(unique(mouse_metadata$celltype)),unique(mouse_metadata$celltype))]
p1 <- DimPlot(object = mouse, group.by = "celltype", reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=TRUE, cols=colsM) + NoLegend() + xlim(xl) + ylim(yl)
p2 <- DimPlot(object = mouse, group.by = "celltype", reduction = "umap", do.return = TRUE, pt.size = 1, 
              label=FALSE, cols=colsM) + NoLegend() + xlim(xl) + ylim(yl)
plot_grid(p1, p2, ncol = 3)
ggsave("UMAP_mouse.pdf",height=6,width=15)
```  
  
Finally, plot SMI-32 (NEFH gene) expression in human and mouse.

```{r plot NEFH, fig.height=5,fig.width=15}
p1 <- FeaturePlot(object = human, reduction = "umap", pt.size = 1.2, features = "NEFH", ncol=4, slot="counts") + xlim(xl) + ylim(yl)
p2 <- FeaturePlot(object = mouse, reduction = "umap", pt.size = 1.2, features = "NEFH", ncol=4, slot="counts") + xlim(xl) + ylim(yl)
plot_grid(p1, p2, ncol = 2)
ggsave("UMAP_NEFH.pdf",height=6,width=12)
```  
  
  
Save UMAP coordinates for human and mouse.  These will be used for mapping Patch-seq data in later scripts.  
  
```{r save umap}
umap_human <- p1$data[,c("UMAP_1","UMAP_2")]
umap_mouse <- p2$data[,c("UMAP_1","UMAP_2")]
save(umap_human,umap_mouse,file="umap_coordinates.RData")
```
  
  
### Assess cluster heterogeneity in mouse and human  
  
To assess cluster heterogeneity, we will compare the variance explained from the first PC in real data against the variance epxlained from the first PC in random shufflings of the data.  Clusters with high heterogeneity should also have at least one signficant PC relative to the expected PC value from chance.  We will downsample to the same number of cells per cluster and use the same number of variable genes calculated the same way in mouse and human for a fair comparison.  We are also excluding outlier samples from the PC calculation to insure that the results are not driven by a small number of outlier or low quality cells, and are excluding sex and mitochondrial genes.  
  
First some housekeeping things (functions, common and non-excluded genes).  
  
```{r define required functions}
## Select the genes of common mouse/human genes that are not sex or mitochondrial genes for later
data(mito_genes)  
data(sex_genes)  
excludeGenes <- sort(unique(c(sex_genes,mito_genes)))  
kpGenes <- intersect(rownames(datVISp),rownames(datMTG))
kpGenes <- sort(setdiff(kpGenes,excludeGenes))

## Required functions
findFromGroupsVector <- function(x,y,...) { 
 x = as.numeric(x)
 x = cbind(x,x)
 f = findFromGroups(x,y,...)
 return(colMeans(f))
}

# This function calculates the top PCs after removing any outlier cells 
# (to account for potential low-quality or misassigned cells)
prcompAfterOutlierRemoval <- function(tnorm.dat,gnUse=1:dim(tnorm.dat)[2],kpCl=1:dim(tnorm.dat)[2],zThresh){
  pcaUse <- prcomp(tnorm.dat[kpCl,gnUse])
  go = TRUE
  # Check for outliers and recalculate PCs until no outliers remain.
  while(go&length(kpCl>10)){
    z1 = pcaUse$x[,1];  z1 = (z1-mean(z1))/sd(z1)
    z2 = pcaUse$x[,2];  z2 = (z2-mean(z2))/sd(z2)
	# Check for outliers
    kpTmpS = pmax(abs(z1),abs(z2))<zThresh
	if (sum(!kpTmpS)==0){
	  go = FALSE
	} else {
      kpCl[kpCl][!kpTmpS] = FALSE
      pcaUse <- prcomp(tnorm.dat[kpCl,gnUse])
    } 
  }
  return(pcaUse)
}
```
  
  
Next, find the actual and permuted PCs in human.  
  
```{r human scramble PCs}
set.seed(1)
nCells  <- 80
nGenes  <- nCells
rCount  <- 10 # 2
pCount  <- 100
zThresh <- 3
pcMax   <- 10
p1h <- p1hr <- list()
clustsH <- sort(unique(annoMTG$cluster_label))[c(3,2,1,4,5)] # Order superficial to deep
for(cls in clustsH){
  print(cls)
  for (i in 1:rCount){
    if(i==1) {
      p1h[[cls]]  <- NULL
      p1hr[[cls]] <- NULL
    }
    set.seed(i)
    kpCell <- sample(colnames(datMTG)[annoMTG$cluster_label==cls],nCells)
    nbt    <- CreateSeuratObject(datMTG[kpGenes,kpCell])
    nbt    <- FindVariableFeatures(nbt, selection.method = "vst", nfeatures = nGenes, mean.cutoff = c(1,Inf), verbose = FALSE)
    datTmp <- as.matrix(FetchData(nbt,vars = VariableFeatures(nbt), slot="counts"))
    pcaUse <- prcompAfterOutlierRemoval(datTmp,zThresh=zThresh)
    varExp <- ((10000*(pcaUse$sdev)^2 / sum(pcaUse$sdev^2))/100)[1:pcMax]
    p1h[[cls]] <- rbind(p1h[[cls]],varExp)
    for (j in 1:pCount){
      set.seed(i*pCount+j)
      datRand <- t(apply(datTmp,1,function(x) return(x[sample(1:length(x),length(x))])))
      colnames(datRand) <- colnames(datTmp)
	    pcaRand <- prcompAfterOutlierRemoval(datRand,zThresh=zThresh)
	    varExpR <- ((10000*(pcaRand$sdev)^2 / sum(pcaRand$sdev^2))/100)[1:pcMax]
	    p1hr[[cls]] <- rbind(p1hr[[cls]],varExp/varExpR)
    }
  }
}
```
  
  
Repeat, finding the actual and permuted PCs in mouse.  
  
```{r mouse scramble PCs}
clustsM <- sort(unique(annoVISp$cluster_label))[c(1,3,2)] # Order superficial to deep
for(cls in clustsM){
  print(cls)
  for (i in 1:rCount){
    if(i==1) {
      p1h[[cls]]  <- NULL
      p1hr[[cls]] <- NULL
    }
    set.seed(i)
    kpCell <- sample(colnames(datVISp)[annoVISp$cluster_label==cls],nCells)
    nbt    <- CreateSeuratObject(datVISp[kpGenes,kpCell])
    nbt    <- FindVariableFeatures(nbt, selection.method = "vst", nfeatures = nGenes, mean.cutoff = c(1,Inf), verbose = FALSE)
    datTmp <- as.matrix(FetchData(nbt,vars = VariableFeatures(nbt), slot="counts"))
    pcaUse <- prcompAfterOutlierRemoval(datTmp,zThresh=zThresh)
    varExp <- ((10000*(pcaUse$sdev)^2 / sum(pcaUse$sdev^2))/100)[1:pcMax]
    p1h[[cls]] <- rbind(p1h[[cls]],varExp)
    for (j in 1:pCount){
      set.seed(i*pCount+j)
      datRand <- t(apply(datTmp,1,function(x) return(x[sample(1:length(x),length(x))])))
      colnames(datRand) <- colnames(datTmp)
	    pcaRand <- prcompAfterOutlierRemoval(datRand,zThresh=zThresh)
	    varExpR <- ((10000*(pcaRand$sdev)^2 / sum(pcaRand$sdev^2))/100)[1:pcMax]
	    p1hr[[cls]] <- rbind(p1hr[[cls]],varExp/varExpR)
    }
  }
}
```  
  
  
Now plot the results, comparing variation across all eight types.  
  
```{r plot heterogeneity results, fig.height=7,fig.width=7}
meanDiff <- sdDiff <- NULL
for (cls in names(p1hr)){
  meanDiff <- rbind(meanDiff,colMeans(p1hr[[cls]]))
  sdDiff   <- rbind(sdDiff, apply(p1hr[[cls]],2,sd))
}
rownames(meanDiff) <- rownames(sdDiff) <- names(p1hr)
colnames(meanDiff) <- colnames(sdDiff) <- paste0("PC",1:pcMax)

# Data
data <- data.frame(
  name=factor(c(spl(clustsH),spl2(clustsM)),levels=c(spl(clustsH),spl2(clustsM))),
  value=meanDiff[,1],
  sd=sdDiff[,1]
)

a = setNames(c(annoVISp$cluster_color,annoMTG$cluster_color),c(annoVISp$cluster_label,annoMTG$cluster_label))
data$cols <- as.character(a[rownames(data)])

# line + dot
ggplot(data) +
  geom_bar( aes(x=name, y=value), stat="identity", fill=data$cols, alpha=1)+#0.5) +
  geom_errorbar( aes(x=name, ymin=value-sd, ymax=value+sd), width=0.4, colour="black", alpha=0.9, size=1.3) +
  ylab("Actual/permuted variance explained by PC1") + xlab("") + theme_bw() +
  theme(axis.text.x = element_text(angle = 90), text = element_text(size=16))
ggsave("heterogeneity.pdf",height=8,width=4)
```
  
  
## Cluster discreteness (based on number of pairwise DEX genes)
  
Using `scrattch.hi`, calculating the number of DEX genes between each pair of clusters, subsampling to 80 cells per cluster and repeating averaging across 10 iterations again.  
  
```{r dex genes using hicat}
de.param   <- de_param(low.th = 1, padj.th = 0.01, lfc.th = 1,    # Recommended parameters for SMART-seq data
                       q1.th = 0.5, q.diff.th = 0.7, de.score.th = 150)
humanCount <- mouseCount <- 0
numIters   <- 10

for (n in 1:numIters){
  ## Human
  cl <- setNames(make.names(annoMTG$cluster_label),annoMTG$sample_id)
  nm <- sort(unique(cl))
  kpSamp <- subsampleCells(cl,subSamp=nCells,seed=n)
  human_de <- de_score(datMTG[kpGenes, kpSamp], cl[kpSamp], de.param)
  human_count <- matrix(0, ncol = length(nm), nrow = length(nm))
  rownames(human_count) <- colnames(human_count) <- nm
  for (i in 1:(length(nm)-1)) for (j in (i+1):length(nm)){
    human_count[i,j] = human_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$up.num
    human_count[j,i] = human_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$down.num
  }
  humanCount = human_count + humanCount
  
  ## Mouse
  cl <- setNames(make.names(annoVISp$cluster_label),annoVISp$sample_id)
  nm <- sort(unique(cl))
  kpSamp <- subsampleCells(cl,subSamp=nCells,seed=n)
  mouse_de <- de_score(datVISp[kpGenes, kpSamp], cl[kpSamp], de.param)
  mouse_count <- matrix(0, ncol = length(nm), nrow = length(nm))
  rownames(mouse_count) <- colnames(mouse_count) <- nm
  for (i in 1:(length(nm)-1)) for (j in (i+1):length(nm)){
    mouse_count[i,j] = mouse_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$up.num
    mouse_count[j,i] = mouse_de$de.genes[[paste(nm[i],nm[j],sep="_")]]$down.num
  }
  mouseCount = mouse_count + mouseCount
}
mouseCount = (mouseCount+t(mouseCount))/numIters
humanCount = (humanCount+t(humanCount))/numIters
print(paste("Human:",sum(humanCount[1:3,1:3])/6))
print(paste("Mouse:",sum(mouseCount[1:3,1:3])/6))
```
  
  
Comparable DEX genes in matched L2/3 types in mouse and human (maybe fewer in human) after matching for number of cells.  Plot these results.  
  
```{r plot dex genes}
out <- t(cbind(humanCount[1:3,],mouseCount))
colnames(out) <- 1:3
out[out==0]=NA
out <- as.data.frame(out)
out$name <- rownames(out) <- data$name

barplot(apply(out[,1:3],1,mean,na.rm=TRUE),las=2,col=data$cols, ylab="# DEX genes wrt matched human/mouse types",border="white")

pdf("discreteness.pdf",height=8,width=4)
barplot(apply(out[,1:3],1,mean,na.rm=TRUE),las=2,col=data$cols, ylab="# DEX genes wrt matched human/mouse types",border="white")
dev.off()
```
  
  
Output session information.  
  
```{r sessionInfo}
sessionInfo()
```      
  